A Comparative Quality Metric for Fuzzing with Logic State Coverage

### 2024-03-04 paper/main.tex

  To establish a more reliable quality metric, we first note that fuzzing 
  aims to reduce the number of unknown \emph{abnormal} behaviors by observing
  more \emph{interesting} (i.e., relating to unknown abnormal) behaviors. The
  more interesting behaviors a fuzzer has observed, the stronger guarantee it
  can provide about the absence of unknown abnormal behaviors. This suggests
  that the number of observed interesting behaviors must directly indicate the
  fuzzing quality.

@@

As it is challenging to directly count interesting program behaviors when
fuzzing produces hundreds of them every second, we propose \emph{logic state
coverage} as a feasible proxy metric of interesting behaviors. A logic state is
a set of all satisfied branches during one execution, where logic state coverage
is the \emph{individual count} of all observed distinct logic states during a
fuzzing campaign. Logic states represent more \emph{interesting} behaviors
(i.e., relating to unknown abnormal behaviors) in a finer granularity, meaning
that observing more interesting behaviors is reliably reflected to larger
coverage.

@@

\pp{Formal justification}
%
The approach of untargeted fuzzing can be formally described as follows. Upon
executing a mutated input, let $P(\mathrm{N})$ be the probability of observing a
\underline{\textbf{n}}ew behavior as a result of execution. Then, the
probability of discovering \underline{\textbf{u}}nknown
\underline{\textbf{a}}bnormal behavior $P(\mathrm{UA}) := P(\mathrm{U} \cap
\mathrm{A})$ is
%
\begin{align}
  P(\mathrm{UA}) 
      &= P(\mathrm{N} \cap \mathrm{UA}) \tag{by $\mathrm{U} \subset \mathrm{N}$} \\
      &= P(\mathrm{N}) \times P(\mathrm{UA} \mid \mathrm{N}) \tag{by chain rule}. 
\end{align}

Here, the second term $P(\mathrm{UA} \mid \mathrm{N})$ is the expectation of
revealing an unknown abnormal behavior given a new behavior, which can be
interpreted as the \emph{interestingness} of the mutation base. In other words,
untargeted fuzzing attempts to (i) increase $P(\mathrm{N})$ to investigate more
program behaviors while (ii) keeping the interestingness $P(\mathrm{UA} \mid
\mathrm{N})$ as high as possible.
%
%Notice that executing more inputs is not the same as observing more distinct
%behaviors, as inputs can exhibit the same behavior (e.g., the same
%early-termination).

\pp{Interestingness in practice}
%
As the interestingness $P(\mathrm{UA} \mid \mathrm{N})$ is the expectation of an
\emph{unobserved} behavior, there is no direct way to measure interestingness. 
%
However, the practice of untargeted fuzzing suggests that
the resulting program behavior is unlikely to be \emph{unknown abnormal} if the
mutation base is highly repetitive, as fuzzers commonly incorporate corpus
minimization \cite{todo} and branch hit count buckets \cite{todo}.
%
The rationales are: (i) if a fewer repetition was normal, more repetition is
also likely normal (i.e., unlikely \emph{abnormal}) and (ii) even if it was
abnormal, it may correspond to pre-observed less-repetition counterparts (i.e.,
unlikely \emph{unknown}).

@@

%To formalize this, let $X=n$ denote when a program behavior takes a 
%branch $X$ $n$ times ($n \ge 1$). Then, the interestingness of a program
%behavior when $X=n$ is;
%%
%\begin{equation}
%  P(\mathrm{abnormal} \mid X=n) \propto 1/n.
%\end{equation}

@@

\pp{Logic state and logic space}
%
A \emph{logic state} $l$ is a set of all satisfied branches during execution,
which is essentially the order-ignored version of a branch history. Formally,
given a branch history $\mathcal{T}_b$ of an execution,
%
\begin{equation}
\label{e:ls}
  l := \{b \mid b \in \mathcal{T}_b\},
\end{equation}

@@

\pp{Interestingness: definition vs. convention}
%
Although the \emph{interestingness} has been widely discussed in recent
untargeted fuzzing, it has never been clearly defined. The conventional
understanding is a property of an \emph{input} having \emph{new
execution paths} in its resulting behavior. In comparison, this work
defines interestingness as a property of a \emph{program behavior} indicating 
the \emph{expectation} of being an abnormal behavior.

Despite an apparent difference between them, the definition \emph{subsumes} the
convention because if an input discovered a new path (conventional
interesting), its behavior would have a higher expectation to be abnormal than
the ones only exhibiting previously checked paths (defined interesting). 
%
Still, there are some disagreements between the definition and the convention,
such as when a behavior \emph{contains} another known abnormal behavior without
discovering any new path. The definition may be refined to resolve such
disagreements.

\pp{Completeness of targeted fuzzing}
%
The completeness of untargeted fuzzing hinges on the observation that the more
program behavior it checks, the less abnormal program behaviors are expected to
be left. This roughly translates to $P(\mathrm{abnormal} \mid \mathrm{new})$
being equal throughout a fuzzing campaign.
%, whereby $P(\mathrm{new} \cap
%\mathrm{abnormal})$ then majorly depends on how high $P(\mathrm{new})$ is;
%proportional to the number of observed program behaviors per time.
%
However, targeted fuzzing (e.g., directed or regression fuzzing)
\emph{prioritizes} code that likely leads to abnormal behaviors, which may result
in a larger $P(\mathrm{abnormal} \mid \mathrm{new})$ at the beginning that slowly
decreases over time.

This affects the fundamental observation for completeness because targeted
fuzzers may reveal more abnormal behaviors even after observing fewer program
behaviors.
%
One possible remedy is refining interestingness to incorporate a complete set of
factors that affects $P(\mathrm{abnormal})$ (e.g., freshness of code).  A
further investigation on interestingness is required.

>>>>

  To establish a more reliable quality metric, we first note that fuzzing 
  aims to reduce the number of \emph{unknown abnormal} behaviors by observing
  more \emph{interesting} (i.e., relating to unknown abnormal) behaviors. The
  more interesting behaviors a fuzzer has observed, the stronger guarantee it
  can provide about the absence of unknown abnormal behaviors. This suggests
  that the number of observed interesting behaviors must directly indicate the
  fuzzing quality.

@@

As it is challenging to directly count interesting program behaviors when
fuzzing produces hundreds of them every second, we propose \emph{logic state
coverage} as a feasible proxy metric of interesting behaviors. A logic state is
a set of all satisfied branches during one execution, where logic state coverage
is the \emph{individual count} of all observed distinct logic states during a
fuzzing campaign. Logic states represent \emph{interesting} behaviors (i.e.,
relating to unknown abnormal behaviors) in a finer granularity, meaning that
observing more interesting behaviors is reliably reflected to larger coverage.

@@

\pp{Formal justification}
%
The approach of untargeted fuzzing can be formally described as follows. Upon
executing a mutated input, let $\Pr(\mathrm{N})$ be the probability of observing a
\underline{\textbf{n}}ew behavior as a result of execution. Then, the
probability of discovering \underline{\textbf{u}}nknown
\underline{\textbf{a}}bnormal behavior $\Pr(\mathrm{UA}) := \Pr(\mathrm{U} \cap
\mathrm{A})$ is
%
\begin{align}
  \Pr(\mathrm{UA}) 
      &= \Pr(\mathrm{N} \cap \mathrm{UA}) \tag{by $\mathrm{U} \subset \mathrm{N}$} \\
      &= \Pr(\mathrm{N}) \cdot \Pr(\mathrm{UA} \mid \mathrm{N}) \tag{by chain rule}. 
\end{align}

Here, the second term $\Pr(\mathrm{UA} \mid \mathrm{N})$ is the expectation of
revealing an unknown abnormal behavior given a new behavior, which can be
interpreted as the \emph{interestingness} of the mutation base. In other words,
untargeted fuzzing attempts to (i) increase $\Pr(\mathrm{N})$ to investigate
more program behaviors while (ii) keeping the interestingness $\Pr(\mathrm{UA}
\mid \mathrm{N})$ high.
%
%Notice that executing more inputs is not the same as observing more distinct
%behaviors, as inputs can exhibit the same behavior (e.g., the same
%early-termination).

\pp{Interestingness in practice}
%
As the interestingness $\Pr(\mathrm{UA} \mid \mathrm{N})$ is the expectation of
an \emph{unobserved} behavior, there is no direct way to measure
interestingness. 
%
However, the common usage of corpus minimization \cite{todo} and branch hit
count buckets \cite{todo} in untargeted fuzzing suggests that a program behavior
is less likely to be \emph{unknown abnormal} (i.e., less interesting) if its
mutation base input is highly repetitive.
%in modern untargeted fuzzing.
%
The rationales are: (i) if a fewer repetition was normal,
more repetition is also likely normal and (ii) even if it was abnormal, it is 
not unknown if it corresponds to pre-observed less-repetition counterparts.

To generalize this practice from mutation base inputs to all inputs (i.e., all
their program behaviors), a program behavior can be deemed less interesting if
it include more repetitions.

@@

%To formalize this, let $X=n$ denote when a program behavior takes a 
%branch $X$ $n$ times ($n \ge 1$). Then, the interestingness of a program
%behavior when $X=n$ is;
%%
%\begin{equation}
%  \Pr(\mathrm{abnormal} \mid X=n) \propto 1/n.
%\end{equation}

@@

\pp{Logic state and logic space}
%
A \emph{logic state} $l$ is a set of all satisfied branches during execution,
which is essentially the order-ignored version of a branch history. Formally,
given a branch history $\mathcal{T}_b$ of an execution,
%
\begin{equation}
\label{e:ls}
  l := \{b \mid b \in \mathcal{T}_b\}.
\end{equation}

@@

%\pp{Interestingness: definition vs. convention}
%%
%Although the \emph{interestingness} has been widely discussed in recent
%untargeted fuzzing, it has never been clearly defined. The conventional
%understanding is a property of an \emph{input} having \emph{new
%execution paths} in its resulting behavior. In comparison, this work
%defines interestingness as a property of a \emph{program behavior} indicating 
%the \emph{expectation} of being an abnormal behavior. 
%
%Despite an apparent difference between them, the definition \emph{subsumes} the
%convention because if an input discovered a new path (conventional
%interesting), its behavior would have a higher expectation to be abnormal than
%the ones only exhibiting previously checked paths (defined interesting). 
%%
%Still, there are some disagreements between the definition and the convention,
%such as when a behavior \emph{contains} another known abnormal behavior without
%discovering any new path. The definition may be refined to resolve such
%disagreements.

\pp{Completeness of targeted fuzzing}
%
The completeness of untargeted fuzzing hinges on the observation that the more
program behavior it checks, the less abnormal program behaviors are expected to
be left. This roughly translates to $\Pr(\mathrm{abnormal} \mid \mathrm{new})$
being equal throughout a fuzzing campaign.
%, whereby $\Pr(\mathrm{new} \cap
%\mathrm{abnormal})$ then majorly depends on how high $\Pr(\mathrm{new})$ is;
%proportional to the number of observed program behaviors per time.
%
However, targeted fuzzing (e.g., directed or regression fuzzing)
\emph{prioritizes} code that likely leads to abnormal behaviors, which may result
in a larger $\Pr(\mathrm{abnormal} \mid \mathrm{new})$ at the beginning that slowly
decreases over time.

This affects the fundamental observation for completeness because targeted
fuzzers may reveal more abnormal behaviors even after observing fewer program
behaviors.
%
One possible remedy is refining interestingness to incorporate a complete set of
factors that affects $\Pr(\mathrm{abnormal})$ (e.g., freshness of code).  A
further investigation on interestingness is required.
